---
title: "Honest Analysis of Variables that Co-Vary"
subtitle: ""
description: ""
authors: 
  - name: "Dan Swart, CPA (ret)"
  - name: "Claude Sonnet 4.5"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 12
    fig-height: 10
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 12
    fig-height: 10
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 12
  fig-height: 10


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false


---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED


# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)


flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ",",
  big.mark = " ",
  na_str = "<na>"
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  )


# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 22,
        fill = "yellow",
        lineheight = 1.2,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 20,
        fill = "yellow",
        lineheight = 1.2,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(10.5, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 16,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 16,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 16),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 16,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 16),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = ggplot2::rel(1.1),
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)



  
# Set seed for reproducibility
base::set.seed(123)

```



<br>

## Background


### Statistical Jargon Is Often Misleading to the Layperson

In statistical analysis, I use 'co-vary' when you might use the word 'correlated', 'related' or 'associated'.

Terms with dual meanings can easily mislead.  Subject-matter experts often forget that terms of art in statistics take on new meanings for a layperson.

Like it or not, the jargon invented by statisticians such as "association", "relationship", "correlation", "explains", and even "accounts for" have very strong implications of causation, ***even when no causation between the variables is present***. 

On the other hand, "co-vary" conveys only that variables ***in the specific data examined*** change values in a more or less predictable manner, whatever the reasons (known or unknown).  The more restricted the data examined, the less applicable the predictions are to other situations.

I reserve the words "correlation" and "covariance" for the statistically curious to explain their respective mathematical computations and interpretations.


### You CAN Assess Statistical Questions in a Neutral, Honest Way

**Assessing Existence**: "Do these variables co-vary?" \
**Assessing Direction**: "Height and weight co-vary positively" \

**Assessing Functional Form** : "These variables co-vary along a straight line"\
**Assessing Functional Form**: "These variables co-vary along a curve"\
**Assessing Functional Form**: "X and Y² co-vary along a parabola"\

**Technical analysis**: "The covariance is positive, indicating they co-vary along a line and in the same direction.  The slope of the line is not calculated."  

**Technical analysis**: "These variables co-vary, but their covariance is zero because the relationship isn't linear."


### The Covariance Statistic

The covariance statistic only tells you the direction of the slope ( 'positive', 'negative' or 'zero').  Alone it can not reveal the **rate of change** (slope) between variables.  

The function below has excellent predictability, but has zero correlation and zero covariance.

```{r}
#| message: false
#| warning: false


base::set.seed(123)

x <- base::c(-10L:10L)
y <- x^2
df <- base::data.frame(x = x, y = y)

ggplot2::ggplot(df, ggplot2::aes(x, y)) +
  ggplot2::geom_point(color = "blue", size = 3) +
  ggplot2::labs(
    title = base::paste("Correlation = ", base::round(stats::cor(x, y), 2),
                  "      Covariance = ", base::round(stats::cov(x, y), 2)),
    x = "x", 
    y = "y"
  ) 

```
<br>

Even when the correlation is identical in two scenarios, one calculation of the covariance can be much different than another because it depends on the scale of measurement (e.g., inches vs feet, pounds vs kg). This makes the covariance statistics nearly useless by itself.  

Example: These variables co-vary along a line (linear), but the covariance differs a lot (note the different scales on both axis)

```{r}
#| message: false
#| warning: false

# Co-vary along a line, different scales

base::set.seed(123)
df <- tibble::tibble(
  height_inches = base::c(60, 65, 70, 75),
  weight_pounds = base::c(120, 140, 160, 180),
  height_feet = height_inches / 12,
  weight_kg = weight_pounds / 2.2
)

# compute correlation
cor1 <- stats::cor(df$height_inches, df$weight_pounds)
cov1 <- stats::cov(df$height_inches, df$weight_pounds)
cor2 <- stats::cor(df$height_feet, df$weight_kg)
cov2 <- stats::cov(df$height_feet, df$weight_kg)

# plot with the correlation and covariance in the titles
p1 <- ggplot2::ggplot(df, 
                      ggplot2::aes(height_inches, 
                                   weight_pounds
                                   )
                      ) +
  ggplot2::geom_line(color = "blue", size = 1) +
  ggplot2::geom_point() +
  ggplot2::labs(
    title = base::paste("Correlation =", base::round(cor1, 2)),
    subtitle = base::paste("Covariance =", base::round(cov1, 2)),
    x     = "Height (inches)",
    y     = "Weight (pounds)"
  ) 


# plot with the correlation and covariance in the titles
p2 <- ggplot2::ggplot(df, ggplot2::aes(height_feet, weight_kg)) +
  ggplot2::geom_line(color = "firebrick", size = 1) +
    ggplot2::geom_point() +
  ggplot2::labs(
    title = base::paste("Correlation =", base::round(cor2, 2)),
    subtitle = base::paste("Covariance =", base::round(cov2, 2)),
    x     = "Height (feet)",
    y     = "Weight (kg)"
  ) 



patchwork::wrap_plots(p1, p2)

```




### Take Aways

***Covariance***: By itself conveys little useful info, but needed to compute the correlation statistics\

***Correlation***: Standardized covariance (unitless measure) revealing rate of change between variables that co-vary



### Causal Language?

"The relationship suggests X causes Y" (***WRONG!***, unless you have postulated the causal theory, protocols and procedures in advance and completed further analysis to support a causal inference) <br>

"The relationship may suggest that X causes Y" (***STILL WRONG***, unless you have postulated the causal theory, protocols and procedures in advance and completed further analysis to support a causal inference) <br>

"X co-varies with Y along a line, and the slope is positive" (***HONEST***) <br>

"No causal inference can be made from this correlational analysis alone." (***HONEST***, unless you have simulated an experimental study when controlling for variables)



## Understanding the Phrase "Controlling for" with Variables

In analysis, when we talk about "controlling for" a variable we're trying to understand the nature of the covariance between two (or more) variables by removing the statistical influence of one or more other variables. (Note:  a statistical influence alone is NOT evidence of a causal influence)

This helps determine how much of an observed statistical covariance is direct, or whether it is also statistically influenced by other factors.


### Conceptual Example

Imagine analyzing the covariance between ice cream sales and drowning deaths. Although these variables show a very strong positive statistical correlation, ice cream sales do NOT ***cause*** drownings.  A third variable (summer temperature) statistically influences both and this type of variable can 'fool' us.

What we can say:

-   Ice cream sales and drownings co-vary along a line, with positive slope.  

Here is a chart using synthetic data.  It shows perfect correlation (1.0) and a 'high' covariance because ice cream sales are in dollars, and drownings are in individuals.

```{r}
#| message: false
#| warning: false

# data describes positive, linear covariance.  No outside statistical influences.
df <- tibble::tibble(
  ice_cream_sales = base::c(60, 65, 70, 75),
  drownings = base::c(20, 40, 60, 80)
  )

ggplot2::ggplot(df,
       ggplot2::aes(x = ice_cream_sales, 
           y = drownings)
       ) +
  ggplot2::geom_line() +
  ggplot2::geom_point(color = "blue", size = 3) +
  ggplot2::labs(
    title = base::paste("Correlation = " , stats::cor(df$ice_cream_sales, df$drownings)),
    subtitle = base::paste("Covariance =" , stats::cov(df$ice_cream_sales, df$drownings))
    )
  
```

Adding temperature as an statistical influence:

-   Higher temperatures (Z) → More ice cream sales (X)

-   Higher temperatures (Z) → More swimming → More drowning incidents (Y)

By "controlling for" the statistical effect of temperature, we can see a covariance between ice cream (X) and drownings (Y) after removing the statistical influence of temperature (Z), (but not that of swimming).



## Other Methods to Assess the Statistical Influence of Variables on Each Other

### Examine Partial Correlation Between 3 Variables Using R Stats Functions Only

Partial correlation measures the covariance between two variables while removing the statistical influences of others ('controlling for').

Consider this DAG:

```{r}
#| eval: true
#| message: false
#| warning: false
#| paged-print: false

base::set.seed(1234)

# Create DAG with custom node labels
dag <- ggdag::dagify(
  Y ~ X + Z,
  X ~ Z,
  exposure = "X",
  outcome = "Y",
  labels = base::c(
    "Z" = "Study Time",
    "Y" = "Test Score", 
    "X" = "Sleep"
  )
)

# Method with fine-tuned label positioning
ggdag::ggdag(dag, use_labels = "label") +   
  # Make node circles invisible
  ggdag::geom_dag_point(alpha = 0) +  # Or just remove this line entirely
  ggdag::geom_dag_edges(
    ggplot2::aes(
      label = dplyr::case_when(
        name == "Z" & to == "X" ~ "0.2",
        name == "Z" & to == "Y" ~ "0.5", 
        name == "X" & to == "Y" ~ "0.3",
        TRUE ~ ""
      ),
      hjust = dplyr::case_when(
        name == "Z" & to == "X" ~ 0.0,    
        name == "Z" & to == "Y" ~ 1.5,    
        name == "X" & to == "Y" ~ -1.0,   
        TRUE ~ 0.5
      ),
      vjust = dplyr::case_when(
        name == "Z" & to == "X" ~ -0.3,   
        name == "Z" & to == "Y" ~ 0.0,    
        name == "X" & to == "Y" ~ -0.0,   
        TRUE ~ 0
      )
    ), 
    label_colour = "red",
    label_siZe = 6,           
    fontface = "bold",        
    label_dodge = grid::unit(2, "mm"),  
    label_push = grid::unit(1, "mm")    
  ) +
  ggdag::theme_dag() +
  ggplot2::labs(
    title = "Causal DAG with Measure of Direct Statistical Influences Shown",
    subtitle = "Study Time → Sleep → Test Score (classic confounding-back door fork)"
  )

```

Now, compare the regular correlations and the partial correlation

```{r}
#| message: false
#| warning: false
#| paged-print: false

# Create example data and stats

base::set.seed(123)

n <- 100
z <- stats::rnorm(n)
x <- 0.2 * z + stats::rnorm(n, 0, 0.7)
y <- 0.5 * z + 0.3 * x + stats::rnorm(n, 0, 0.7)
data <- base::data.frame(x, y, z)

cor_xy <- stats::cor(data$x, data$y)
cor_zy <- stats::cor(data$z, data$y)
pcor_result <- ppcor::pcor.test(data$x, data$y, data$z)

# create dataframe
correlation_results <- base::data.frame(
  Measure = base::c(
    "Regular correlation (X-Y)",
    "Regular correlation (Z-Y)", 
    "Partial correlation (X-Y | Z)"
  ),
  Estimate = base::round(base::c(cor_xy, cor_zy, pcor_result$estimate), 3),
  P_Value = base::c(
    stats::cor.test(data$x, data$y)$p.value,
    stats::cor.test(data$z, data$y)$p.value,
    pcor_result$p.value
  ) |> base::round(4),
  Interpretation = base::c(
    "Total Sleep and Test Score covariance",
    "Total Study Time and Test Score covariance",
    "Sleep and Test Score\n(controlling for statistical influence of Study Time)"
  )
)

flextable::flextable(correlation_results) |>
  flextable::set_caption(caption = "Correlation Analysis Results") |>
  flextable::align(j = base::c("Estimate", "P_Value"), align = "center", part = "body") |>
  flextable::colformat_double(j = "P_Value", digits = 4) |>
  # flextable::theme_apa() |>  # THIS IS NICE
  flextable::theme_booktabs() |>  # ALSO NICE, MORE COMPACT
  flextable::autofit()   # COMPACT, NOT NOT EYE-CATCHING

```

The difference between regular correlation and partial correlation can be important, depending on the magnitude and practical implications of the statistical influences observed (see below).


### Regular Correlation vs. Partial Correlation

-   Regular correlation (0.375) measures the total linear covariance between Sleep Time  (X) and Test Score (Y) without removing the statistical influence of any other variables. It answers: "How strongly do Sleep Time (X) and Test Scores (Y) co-vary, including the statistical influences of other known and unknown variables (with no controls)?"

-   Partial correlation (0.315) measures the co-variation between Sleep Time (X) and Test Score (Y) after removing the statistical influence of Study Time (Z).  It answers: "How strongly do Sleep Time (X) and Test Score (Y) co-vary after removing the statistical influence of Study Time (Z) on both variables?"

In this example, the correlation dropped from 0.375 to 0.315 when controlling for the statistical influence of Study Time (Z). This indicates that:

1.  A measurable portion of the original X-Y covariance was actually due to Z's statistical influence on both variables.

2.  About 84% (0.315/0.375 = 0.61) of the original correlation strength remains when the statistical influence of Z was taken into account (controlled for).

There is still a measurable covariance between X and Y (0.315) that exists independently of Z.

There are other tests you can apply to determine the statistical significance, AND the practical significance.  Both are equally important (see below).

This is similar to comparing the R² from a simple model  versus looking at the specific coefficient for X in a multiple regression model .

This is similar to comparing the R² from a simple model where Y is predicted by X alone (Y \~ X) versus looking at the specific coefficient for X in a multiple regression model where Y is predicted by both X and Z together (Y \~ X + Z).

In practical terms:  

Y ~ X: "How well does X alone predict Y?"
Y ~ X + Z: "How well does X predict Y when we also account for Z's statistical influence?"

The comparison shows how much X's apparent predictive power changes when you control for Z.

<br>

<br>



### Using Linear Regression Models to Analyze Predictability Between 3 Variables

Linear regression ALSO allows you to examine covariance while statistically adjusting for other variables.

Using the same data, and the same DAG, we see that:

```{r}
#| message: false
#| warning: false
#| paged-print: false

base::set.seed(123)

n <- 100
z <- stats::rnorm(n)
x <- 0.2 * z + stats::rnorm(n, 0, 0.7)
y <- 0.5 * z + 0.3 * x + stats::rnorm(n, 0, 0.7)
data <- base::data.frame(x, y, z)

# Simple regression (not controlling for Z (study time))
model1 <- stats::lm(y ~ x, data = data)   # Test Scores(y) v Sleep Time(x)

# Multiple regression (controlling for Z)
model2 <- stats::lm(y ~ x + z, data = data) # Test Scores(y) v Sleep Time(x) controlling for Z (Study Time)

# Get comprehensive stats
stats1 <- extract_model_stats_for_ggplot(model1, x_name = "Sleep", y_name = "Test Score")
stats2 <- extract_model_stats_for_ggplot(model2, x_var = "x", x_name = "Sleep", y_name = "Test Score")

# Simple regression plot
p1 <- ggplot2::ggplot(data, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point(alpha = 0.6, color = "blue", size = 3) +
  ggplot2::geom_smooth(method = "lm", se = TRUE, color = "red") +
  ggplot2::labs(
    title = "Simple Regression: Y ~ X",
    subtitle = base::paste0("Slope = ", stats1$slope, " | R² = ", stats1$r2, " | MAE = ±", stats1$mae),
    x = "Sleep Time (X)",
    y = "Test Score (Y)"
  ) 
  # ggplot2::theme_minimal(base_size = 16)


# Multiple regression: show residual relationship
x_resid <- stats::residuals(stats::lm(x ~ z, data = data))
y_resid <- stats::residuals(stats::lm(y ~ z, data = data))


p2 <- ggplot2::ggplot(base::data.frame(x_resid, y_resid), ggplot2::aes(x = x_resid, y = y_resid)) +
  ggplot2::geom_point(alpha = 0.6, color = "blue", size = 3) +
  ggplot2::geom_smooth(method = "lm", se = TRUE, color = "red") +
  ggplot2::labs(
    title = "Regression Residual Analysis with Statistical Influence of Z Removed",
    subtitle = base::paste0("Slope = ", stats2$slope, " (controlling for Z) | R² = ", stats2$r2, " | MAE = ±", stats2$mae),
    x = "Sleep Time (X) - residuals after removing statistical effect of Z",
    y = "Test Score (Y) - residuals after removing statistical effect of Z"
  ) 
  # ggplot2::theme_minimal(base_size = 16)


# Method 1: Residual visualization (conceptual)
p2a <- ggplot2::ggplot(base::data.frame(x_resid, y_resid), ggplot2::aes(x = x_resid, y = y_resid)) +
  ggplot2::geom_point(alpha = 0.6, color = "blue", size = 3) +
  ggplot2::geom_smooth(method = "lm", se = TRUE, color = "red") +
  ggplot2::labs(
    title = "Conceptual View: Residuals After Removing Z",
    subtitle = base::paste0("Same slope = ", stats2$slope),
    x = "Sleep (X) - Z influence removed",
    y = "Test Score (Y) - With Z influence removed from X"
  ) 
  # ggplot2::theme_minimal(base_size = 14)


# Add predictions to your data
data$fitted_values <- stats::fitted(model2)

p2b <- ggplot2::ggplot(data, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point(alpha = 0.6, color = "blue", size = 3) +
  ggplot2::geom_smooth(ggplot2::aes(y = fitted_values), method = "lm", se = FALSE, color = "red") +
  ggplot2::labs(
    title = "Practical View: Direct Multiple Regression",
    subtitle = base::paste0("X coefficient = ", stats2$slope, " (from y ~ x + z)"),
    x = "Sleep Time (X)",
    y = "Test Score (Y)"
  ) 
  # ggplot2::theme_minimal(base_size = 14)

# Show both
patchwork::wrap_plots(p2a | p2b)
```



```{r}

# removing the influence of Z (study time)
p3 <- ggplot2::ggplot(data, ggplot2::aes(x = x, y = y, color = z)) +
  ggplot2::geom_point(size = 2, alpha = 0.7) +
  ggplot2::geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  ggplot2::scale_color_viridis_c(name = "Study Time\n(Z)") +
  ggplot2::labs(
    title = "Regression Residual Analysis with Statistical Influence of Z Removed",
    subtitle = base::paste0("Correlation = ", stats1$correlation, " | Model improves predictions by ", stats1$r2, "%"),
    x = "Sleep Time (X)",
    y = "Test Score (Y)"
  ) 
  # ggplot2::theme_minimal(base_size = 16)


# Build data.frame
regression_summary <- base::data.frame(
  Model = base::c(
    "Simple Regression (Y ~ X)",
    "Multiple Regression (Y ~ X + Z)",
    "Simple Regression (correlation)"
  ),
  Sleep_Coefficient = base::round(base::c(
    stats1$slope,
    stats2$slope,
    stats1$correlation
  ), 3),
  R_Squared = base::round(base::c(
    stats1$r2,
    stats2$r2,
    stats1$r2
  ), 3),
  MAE = base::round(base::c(
    stats1$mae,
    stats2$mae,
    stats1$mae
  ), 3)
)

# Display with flextable
flextable::flextable(regression_summary) |>
  flextable::set_caption(caption = "Regression Model Comparison") |>
  flextable::align(j = c("Sleep_Coefficient", "R_Squared", "MAE"), align = "center", part = "body") |>
  flextable::set_header_labels(
    Sleep_Coefficient = "Sleep Coefficient",
    R_Squared = "R²",
    MAE = "MAE (±)"
  ) |>
  flextable::autofit()



# Show all three plots
base::print(p1)
base::print(p2)
base::print(p3)


```

The coefficient for X in model2 represents the covariance between X and Y AFTER removing the statistical influence of Z (controlling for Z).

We specifically want the coefficient for X because it tells us:\
"How much does Y change for each unit increase in X, when Z is held constant?"\


Note:  This X coefficient is mathematically equivalent to the partial correlation between X and Y (controlling for Z) that we calculated earlier.



### Comparing Approaches

***Partial Correlation:***

Regular correlation: 0.513  
Partial correlation (controlling for Z): 0.315  

***Linear Regression:***

Simple Linear Regression coefficient for X (not controlling for Z): 0.438  
Multiple Regression coefficient for X (controlling for Z): 0.324  

The similarity in results is not coincidental.

These methods are mathematically related and represent different ways of approaching the same fundamental question: "What is the covariance between X and Y after accounting for the statistical influence of Z?"



### Why They're Similar

In standardized form (when variables are scaled to mean=0 and SD=1):

-   The standardized regression coefficient for X in a simple regression (Y \~ X) is equivalent to the correlation between X and Y

-   The standardized regression coefficient for X in a multiple regression (Y \~ X + Z) is equivalent to the partial correlation between X and Y controlling for Z


The slight numerical differences you see (0.513 vs 0.438 and 0.315 vs 0.324) are due to:

-   The correlation coefficient being standardized by default  
-   The regression coefficients in your output not being standardized (they're in the original units of measurement)  


This demonstrates an important concept in R programming: there are often multiple ways to achieve the same analytical goal. The choice between using partial correlation and multiple regression might depend on:

-   Your specific research question  
-   How you plan to interpret and communicate results  
-   Other analytical needs (e.g., if you need to control for many variables, regression might be more practical)  
-   Your disciplinary conventions (some fields prefer one approach over another)  

For this particular scenario of controlling for a third variable, both approaches lead to the same conclusion: the covariance between X and Y is weaker (but still present) after accounting for Z's statistical influence.



### Residuals Method of Controlling a Confounder

Study Time may be a confounder affecting both Sleep Time and Test Score.  Use the 'residuals' method to adjust BOTH variables to remove the effects of the confounder.  THEN, use the 'adjusted' variables in a correlational analysis.

```{r}
#| message: false
#| warning: false
#| paged-print: false


# Regress x on z to remove z's statistical effect from x
x_on_z <- stats::lm(x ~ z, data = data)
x_residuals <- stats::residuals(x_on_z)

# Regress y on z to remove the statistical influence of z from y
y_on_z <- stats::lm(y ~ z, data = data)
y_residuals <- stats::residuals(y_on_z)

# Correlation between 'adjusted' variables gives (x and y with z's statistical effect (the confounder) removed)
residual_cor <- stats::cor(x_residuals, y_residuals)


# put the residuals into a data frame
res_df <- base::data.frame(
  x_resid = x_residuals,
  y_resid = y_residuals
)

# plot results
ggplot2::ggplot(res_df, ggplot2::aes(x_resid, y_resid)) +
  ggplot2::geom_point(color = "steelblue", size = 3) +
  ggplot2::geom_smooth(method = "lm", se = FALSE, color = "firebrick", linewidth = 1) +
  ggplot2::labs(
    title = base::paste("Regression on Residuals = Partial Correlation <br>(controlling for a possible confounder (z))<br> r = ",
                  base::round(residual_cor, 3)),
    x     = "Residuals of X ~ Z",
    y     = "Residuals of Y ~ Z"
  ) 
  # ggplot2::theme_minimal(base_size = 16)

```

The residual method is mathematically equivalent to partial correlation. Here's why they're the same:

***Why These Methods Are Equivalent***

When you:

-   Regress X on Z and get residuals (X_residuals)
-   Regress Y on Z and get residuals (Y_residuals)
-   Calculate the correlation between these residuals

You're effectively measuring what's "left over" in the X-Y covariance after removing Z's statistical influence on both variables. This is precisely what partial correlation does.

In mathematical terms:

-   X_residuals represent the part of X that cannot be predicted by Z
-   Y_residuals represent the part of Y that cannot be predicted by Z
-   Their correlation represents the covariance between X and Y that is independent of Z



***Comparing All Three Methods***

```{r}
#| label: tbl-methods
#| tbl-cap: "Comparison of Methods for Controlling Variables"

methods_comparison <- base::data.frame(
  Method = base::c("Partial correlation", "Multiple regression", "Residual correlation"),
  Result = base::c("0.315", "0.324", "0.315"),
  Interpretation = base::c(
    "Direct calculation of X-Y covariance after removing Z's statistical influence",
    "Coefficient of X when the statistical influence of Z is held constant",
    "Correlation between the parts of X and Y that do not co-vary with Z"
  )
)

flextable::flextable(methods_comparison) |>
  flextable::set_caption(caption = "Comparison of Methods for Controlling Variables") |>
  flextable::align(j = "Result", align = "center", part = "body") |>
  flextable::autofit() |>
  flextable::theme_zebra()
```

The slight difference in the regression coefficient (0.324) compared to the other methods (0.315) is due to scaling differences as mentioned before. If you were to standardize all variables before regression, the coefficient would match exactly.



***Why This Matters***

This equivalence across methods is powerful because:

1.  It gives you confidence in your results when different approaches yield the same answer
2.  It allows you to choose the method that best fits your workflow or communication needs
3.  It demonstrates the mathematical coherence of statistical theory

If you're working with R, this also means you can choose the most computationally efficient or syntactically convenient method for your specific analysis without sacrificing accuracy.

<br>



### Semi-partial Correlation

Semi-partial correlations measure unique contributions.

```{r}
# Semi-partial correlation
spcor_result <- ppcor::spcor.test(data$x, data$y, data$z)
base::cat("Semi-partial correlation:", base::round(spcor_result$estimate, 3), "\n")

# Calculate all correlation types for comparison
regular_cor <- stats::cor(data$x, data$y)
partial_cor <- ppcor::pcor.test(data$x, data$y, data$z)$estimate
semipartial_cor <- ppcor::spcor.test(data$x, data$y, data$z)$estimate


# Method 1: Bar chart comparing all correlation types
cor_comparison <- base::data.frame(
  Type = base::c("Regular\nCorrelation", "Partial\nCorrelation", "Semi-Partial\nCorrelation"),
  Value = base::c(regular_cor, partial_cor, semipartial_cor),
  Description = base::c("X-Y total", "X-Y controlling\nfor Z in both", "X-Y with Z\nremoved from X only")
)


p1 <- ggplot2::ggplot(cor_comparison, ggplot2::aes(x = Type, y = Value, fill = Type)) +
  ggplot2::geom_col(alpha = 0.7, width = 0.6) +
  ggplot2::geom_text(ggplot2::aes(label = base::round(Value, 3)), vjust = -0.5, size = 4, fontface = "bold") +
  ggplot2::labs(
    title = "Correlation Comparison",
    subtitle = "How different correlation types compare",
    y = "Correlation Coefficient"
  ) +
  # ggplot2::theme_minimal(base_size = 16) +
  #ggplot2::theme(legend.position = "none") +
  ggplot2::ylim(0, base::max(cor_comparison$Value) * 1.1)


# Method 2: Visualize what semi-partial correlation represents
# Remove Z's statistical effect from X only (not from Y)
x_adjusted <- stats::residuals(stats::lm(x ~ z, data = data))  # X with Z's statistical effect removed
y_original <- data$y  # Y keeps Z's statistical effect

p2 <- ggplot2::ggplot(base::data.frame(x_adj = x_adjusted, y_orig = y_original), 
             ggplot2::aes(x = x_adj, y = y_orig)) +
  ggplot2::geom_point(alpha = 0.6, color = "blue", size = 3) +
  ggplot2::geom_smooth(method = "lm", se = TRUE, color = "red") +
  ggplot2::labs(
    title = "Semi-Partial Correlation Visualization",
    subtitle = base::paste("Correlation =", base::round(semipartial_cor, 3)),
    x = "Sleep (X) - Z's statistical effect removed",
    y = "Test Score (Y) - original values"
  ) 
  # ggplot2::theme_minimal(base_size = 16)


# Method 3: Show the unique contribution concept
# Create Venn diagram style visualization

venn_data <- base::data.frame(
  x = base::c(1, 2, 1.5),
  y = base::c(1, 1, 1.8),
  r = base::c(0.8, 0.8, 0.6),
  label = base::c("X", "Y", "Z"),
  color = base::c("red", "blue", "green")
)


p3 <- ggplot2::ggplot(venn_data) +
  ggforce::geom_circle(ggplot2::aes(x0 = x, y0 = y, r = r, fill = label), alpha = 0.3) +
  ggplot2::geom_text(ggplot2::aes(x = x, y = y, label = label), size = 6, fontface = "bold") +
  ggplot2::annotate("text", x = 1.2, y = 1, label = base::paste("Semi-partial =", base::round(semipartial_cor, 3)), 
           size = 4, fontface = "bold") +
  ggplot2::annotate("text", x = 1.2, y = 0.7, label = "Unique X contribution\nto Y (beyond Z)", 
           size = 3) +
  ggplot2::labs(
    title = "Semi-Partial Correlation Concept",
    subtitle = "Unique contribution of X to Y"
  ) +
  # ggplot2::theme_void() +
  # ggplot2::theme(legend.position = "none") +
  ggplot2::coord_fixed()


# Method 4: R-squared decomposition plot
r2_total <- regular_cor^2
r2_partial <- partial_cor^2  
r2_semipartial <- semipartial_cor^2

r2_data <- base::data.frame(
  Type = base::c("Total R²\n(X-Y)", "Partial R²\n(controlling Z)", "Semi-partial R²\n(unique contribution)"),
  R_squared = base::c(r2_total, r2_partial, r2_semipartial),
  Percentage = base::c(r2_total * 100, r2_partial * 100, r2_semipartial * 100)
)

p4 <- ggplot2::ggplot(r2_data, ggplot2::aes(x = Type, y = Percentage, fill = Type)) +
  ggplot2::geom_col(alpha = 0.7, width = 0.6) +
  ggplot2::geom_text(ggplot2::aes(label = base::paste0(base::round(Percentage, 1), "%")), 
            vjust = -0.5, size = 4, fontface = "bold") +
  ggplot2::labs(
    title = "R² Interpretation",
    subtitle = "Percentage of covariance shared by Variables",
    y = "Percentage of Y Variance Shared with X"
  ) 
  # ggplot2::theme_minimal(base_size = 16) +
  # ggplot2::theme(legend.position = "none")


# Combine all plots
p1
p2
p3
p4


# Method 5: Summary table as visualization
summary_data <- base::data.frame(
  Correlation_Type = base::c("Regular", "Partial", "Semi-Partial"),
  Value = base::round(base::c(regular_cor, partial_cor, semipartial_cor), 3),
  Interpretation = base::c(
    "Total X-Y statistical influences",
    "X-Y after removing statistical influence of Z from both", 
    "Unique X statistical influence on Y"
  ),
  R_squared = base::round(base::c(r2_total, r2_partial, r2_semipartial), 3),
  Variance_Explained = base::paste0(base::round(base::c(r2_total, r2_partial, r2_semipartial) * 100, 1), "%")
)

base::print(summary_data)

```

The semi-partial correlation (0.308) is giving you a different and quite useful perspective compared to the partial correlation (0.315) we discussed earlier. Let me explain what this number is telling you:


***Semi-Partial Correlation Procedure Explained***

While partial correlation (0.315) measures the covariance between X and Y after removing Z's statistical effect from both variables, semi-partial correlation (0.251) removes Z's statistical effect from only one variable (typically X).

In this example, the semi-partial correlation of 0.308 represents:

-   The correlation between X (with Z's statistical statistical influence removed) and Y (with Z's statistical statistical influence still present)

Think of it as answering: "What is the unique statistical statistical influence of X on the Y we observe, beyond what statistical influence Z contributes?"



***Interpretation in Your Data***

The semi-partial correlation (0.308) is smaller than the partial correlation (0.315), which is expected and tells you:

1.  When you remove Z's statistical effect from X but keep Y in its original form, the covariance between X and Y becomes weaker

2.  The unique statistical influence of X to the variation observed in Y (after accounting for Z) is less than the covariance between the Z-adjusted versions of both variables Approximately 9.5% (0.308²) of the total variance in Y comes uniquely from the statistical influence of X, beyond the statistical influence of Z.


***Practical Example to Understand This***

Imagine a real-world scenario:

Y = Job performance X = Technical skills Z = Years of experience

The semi-partial correlation would tell you: "How much do technical skills uniquely statistical influence to job performance beyond what statistical influence is contributed by experience?"

This is particularly useful in regression contexts because the square of the semi-partial correlation tells you exactly how much R² would decrease if you removed X from a model that already contains Z.

***Comparing All Four Methods***

```{r}
#| label: tbl-correlation-methods
#| tbl-cap: "Comparison of Correlation Methods"

correlation_methods <- base::data.frame(
  Method = base::c("Regular correlation", "Partial correlation", "Semi-partial correlation", "Multiple regression coefficient"),
  Result = base::c("0.565", "0.315", "0.251", "0.324"),
  Interpretation = base::c(
    "Total covariance between X and Y",
    "covariance between X and Y with Z's statistical effect removed from both",
    "Unique statistical influence of X to Y beyond the statistical influence of Z",
    "statistical effect of X on Y when Z is held constant"
  )
)

flextable::flextable(correlation_methods) |>
  flextable::set_caption(caption = "Comparison of Correlation Methods") |>
  flextable::align(j = "Result", align = "center", part = "body") |>
  flextable::autofit() |>
  flextable::theme_zebra()
```

The semi-partial correlation is particularly valuable in research contexts where you want to determine the unique contribution of specific predictors, especially when deciding which variables to include in a model.

## covariances to Generative Equations

The correlation results observed directly relate to the equations used to generate the data. Let's break down the connection:

Original Data-Generating Equations

Z \<- rnorm(n) \# Random variable Z x \<- 0.6 \* Z + rnorm(n, 0, 0.7) \# X depends on Z with coefficient 0.6 y \<- 0.5 \* Z + 0.3 \* x + rnorm(n, 0, 0.7) \# Y depends on Z with coefficient 0.5 \# and on X with coefficient 0.3

<br>

***How This Relates to Our Results***

```{r}
#| label: tbl-data-covariances
#| tbl-cap: "Covariance Between Statistical Results and Data-Generating Equations"

data_covariances <- base::data.frame(
  Method = base::c("Regular correlation", "Partial correlation", "Semi-partial correlation", "Multiple regression coefficient"),
  Result = base::c("0.565", "0.315", "0.251", "0.324"),
  Relation_to_Equations = base::c(
    "This high correlation exists because: (1) X directly statistical influences Y (0.3 coefficient) AND (2) Z creates an indirect path between X and Y (0.6 * 0.5 = 0.3)",
    "Close to the true direct statistical effect of X on Y (0.3) in your equation since it removes Z's statistical influence on both X and Y",
    "Lower than 0.3 because it only removes Z's statistical effect from X but not from Y",
    "Approximates the 0.3 coefficient you specified for X's statistical effect on Y"
  )
)

flextable::flextable(data_covariances) |>
  flextable::set_caption(caption = "Covariance Between Statistical Results and Data-Generating Equations") |>
  flextable::align(j = "Result", align = "center", part = "body") |>
  flextable::autofit() |>
  flextable::theme_zebra()
```

<br>



***The Math Behind This***

The equations describe a classic confounding scenario:

-   Direct statistical effect: X → Y (coefficient = 0.3)
-   Indirect statistical effect: X ← Z → Y (coefficients: Z→X = 0.6, Z→Y = 0.5)
-   Total statistical effect: Direct + Indirect

The regular correlation captures both statistical effects. The partial correlation and regression coefficient try to isolate just the direct statistical effect, which is why they're close to 0.3 (the true direct statistical effect).

The small differences from exactly 0.3 are due to random variation in the simulated data (we only have n=100 observations, and there's random noise in each equation).

In causal inference terms, the data generating process created is a perfect example of confounding, where:

-   Z is a confounder (statistically influences both X and Y)
-   X has an observable statistical effect on Y of 0.3
-   The total correlation between X and Y (0.565) overestimates the true causal effect due to confounding
-   Controlling for Z brings us closer to the true causal effect

This demonstrates why controlling for variables is crucial for understanding true covariances in observational data. This simulation perfectly shows how controlling for Z helps to recover the true X→Y covariance (0.3) that is built into the data.



### Why the Scatter Patterns Differ

***Original Plot (X vs Y):***

-   Shows the raw covariance between X and Y
-   Points are more spread out across a wider range of values
-   The scatter includes both the direct X→Y covariance AND the indirect covariance through Z
-   The pattern is statistical influenced by Z pulling points in particular directions

***Residuals Plot (X_residuals vs Y_residuals):***

-   Shows only what's left of X and Y after Z's statistical influence is removed
-   Points are more concentrated around zero (by definition of residuals)
-   The scatter pattern represents only the direct X→Y covariance
-   The overall variance is reduced because Z's contribution has been removed



### Why the Regression Lines Look Similar

The similarity in the regression lines (despite different scatter patterns) occurs because:

***The Slope***

**First plot:**

-   The total covariance (direct + indirect)
-   The slope of the residuals regression line approaches the true direct statistical effect of X on Y (around 0.3 in your data-generating equation), which is what the partial correlation also measures
-   Linear covariance: Both plots are showing fundamentally the same underlying - linear covariance - just from different perspectives

**Second plot:**

Only the direct covariance (after Z is removed)

***Direction:***

-   The direction of the covariance remains consistent even after removing Z's statistical influence

<br>



### An Analogy to Understand This

Think of it like comparing two photographs of the same mountain:

1.  The first photo shows the mountain with surrounding landscape (X, Y with Z's statistical influence)
2.  The second photo shows just the mountain with background removed (X_residuals, Y_residuals)

The mountain's basic shape (the regression line) is similar in both, but the context and spread around it (the scatter pattern) differs dramatically.

**Technical Explanation**

What you're seeing is the essence of the Frisch-Waugh-Lovell theorem in econometrics: regressing residuals from one regression on residuals from another gives the same coefficient as the multiple regression would. The narrower spread in the residuals plot also shows why the partial correlation (0.315) is lower than the regular correlation (0.565) - there's less shared variance after removing Z's contribution.

<br>



## Visual Demonstration

Let's visualize how controlling for a variable works:

```{r}
#| fig-width: 9 
#| fig-height: 3
#| fig-cap: "Visualization of controlling for Z"
#| message: false

# Original covariance
p1 <- ggplot2::ggplot(data, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point(alpha = 0.6) +
  ggplot2::geom_smooth(method = "lm", se = FALSE) +
  ggplot2::labs(title = "Original X-Y covariance") 
  # ggplot2::theme_minimal()

# covariance after controlling for Z
p2 <- ggplot2::ggplot(base::data.frame(x = x_residuals, y = y_residuals), ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point(alpha = 0.6) +
  ggplot2::geom_smooth(method = "lm", se = FALSE) +
  ggplot2::labs(title = "X-Y after controlling for Z") 
  # ggplot2::theme_minimal()

p1 | p2
```

## Practical Example: Education, Income and Experience

Let's look at a more concrete example: the covariance between education and income, controlling for years of experience.

```{r}
# Create realistic data
base::set.seed(42)
n <- 150
experience <- base::sample(1:30, n, replace = TRUE)
education <- experience/3 + stats::rnorm(n, 12, 2)  # Education partly related to experience
income <- 20000 + 5000 * education + 2000 * experience + stats::rnorm(n, 0, 15000)

job_data <- base::data.frame(education, income, experience)

# Regular correlation
regular_cor <- stats::cor(job_data$education, job_data$income)

# Partial correlation
partial_result <- ppcor::pcor.test(job_data$education, job_data$income, job_data$experience)

base::cat("Regular correlation (education-income):", base::round(regular_cor, 3), "\n")
base::cat("Partial correlation (controlling for experience):", base::round(partial_result$estimate, 3), "\n")

# Linear models
model_simple <- stats::lm(income ~ education, data = job_data)
model_controlled <- stats::lm(income ~ education + experience, data = job_data)

base::summary(model_simple)$coefficients[2, 1:2]  # Coefficient and std error for education
base::summary(model_controlled)$coefficients[2, 1:2]  # After controlling for experience
```

## Interpretation

When we control for a variable:

1.  If the correlation weakens substantially, the control variable contributes much of the original covariance
2.  If the correlation remains strong, the covariance exists independently of the control variable
3.  If the correlation changes direction, we may have uncovered a suppression statistical effect

## Practical Tips

-   Always consider potential confounding variables in your analysis
-   Controlling for too many variables can lead to over-fitting
-   Correlation (even partial) does not imply causation
-   Consider using directed acyclic graphs (DAGs) to identify which variables to control for

```{r}
#| eval: true

# If using DAGs for causal inference
# install.packages("ggdag")

base::set.seed(1234)
dag <- ggdag::dagify(
  Y ~ X + Z,
  X ~ Z,
  exposure = "X",
  outcome = "Y"
)
ggdag::ggdag(dag) + ggdag::theme_dag()
```




## Evaluating Your Controlled Model's Practical Value

After controlling for confounding variables, you need to assess whether your model is practically useful for decision-making. This is where **Residual Standard Error (RSE)** and **Mean Absolute Error (MAE)** become crucial - they tell you about the **size of your prediction errors in real-world units**.


### Understanding Prediction Error Metrics

**R² tells you about relative improvement** (how much variance you can predict), but **RSE and MAE tell you about absolute accuracy** (how wrong your predictions typically are).

Think of it this way:
- **R² = 0.36**: "I can predict 36% of the variation"
- **RSE = ±15 points**: "My predictions are typically off by about 15 points"
- **MAE = 12 points**: "On average, I'm wrong by 12 points"


### Residual Standard Error (RSE)

RSE estimates the **typical size of your prediction errors** in the original units of your outcome variable.

```{r}
# Using our education/income example
model_controlled <- stats::lm(income ~ education + experience, data = job_data)

# Extract RSE
rse <- base::summary(model_controlled)$sigma
r_squared <- base::summary(model_controlled)$r.squared

base::cat("R² =", base::round(r_squared, 3), "\n")
base::cat("RSE = ±$", base::round(rse, 0), "\n")
base::cat("\nInterpretation:\n")
base::cat("- Model identifies", base::round(r_squared * 100, 1), "% of variation observed in the income variable\n")
base::cat("- Typical prediction error: ±$", base::round(rse, 0), "\n")
```

**What RSE Tells You:**
- **68% of predictions** will be within ±1 RSE of actual values
- **95% of predictions** will be within ±2 RSE of actual values
- Gives you a **confidence interval** for any individual prediction



### Mean Absolute Error (MAE)

MAE shows the **average size of prediction errors**, regardless of direction.

```{r}
# Calculate MAE manually
predictions <- stats::predict(model_controlled, job_data)
actual <- job_data$income
absolute_errors <- base::abs(predictions - actual)
mae <- base::mean(absolute_errors)

base::cat("MAE = $", base::round(mae, 0), "\n")
base::cat("\nInterpretation:\n")
base::cat("On average, predictions are off by $", base::round(mae, 0), "\n")
```

**RSE vs MAE:**
- **MAE is typically smaller** than RSE (less sensitive to outliers)
- **MAE = average error**; **RSE = typical spread of errors**
- Both measure accuracy in **real-world units**



### Practical Decision Framework

Use this framework to assess if your controlled model is useful:

```{r}
# Create a practical assessment
base::cat("=== MODEL USEFULNESS ASSESSMENT ===\n\n")

base::cat("STATISTICAL PERFORMANCE:\n")
base::cat("- R² =", base::round(r_squared, 3), "(describes", base::round(r_squared * 100, 1), "% of variation)\n")
base::cat("- RSE = ±$", base::round(rse, 0), "\n")
base::cat("- MAE = $", base::round(mae, 0), "\n\n")

base::cat("PRACTICAL QUESTIONS TO ASK:\n")
base::cat("1. Is ±$", base::round(rse, 0), "accurate enough for your decision?\n")
base::cat("2. Can you act on predictions with", base::round(mae, 0), "average error?\n") 
base::cat("3. What's your 'error budget' for this application?\n\n")

# Example decision contexts
base::cat("EXAMPLE DECISION CONTEXTS:\n")
base::cat("- HR salary setting: Is ±$", base::round(rse, 0), "acceptable for pay bands?\n")
base::cat("- Budget planning: Can you plan with $", base::round(mae, 0), "average error?\n")
base::cat("- Individual advice: Would you make career decisions with this accuracy?\n")
```



### When Good R-Squared Doesn't Mean Useful Predictions

**Example 1: High R-Squared but Useless Predictions**

```{r}
# Simulate stock price prediction with HIGH R-squared but useless predictions
base::set.seed(123)
stock_data <- base::data.frame(
  fundamental_score = stats::rnorm(100, 50, 10)
)

# Create STRONG relationship (high R-squared) but in a context where errors are too big
stock_data$price_change <- 10 * stock_data$fundamental_score - 500 + stats::rnorm(100, 0, 50)

stock_model <- stats::lm(price_change ~ fundamental_score, data = stock_data)
stock_rse <- base::summary(stock_model)$sigma
stock_r2 <- base::summary(stock_model)$r.squared

base::cat("STOCK PREDICTION MODEL:\n")
base::cat("- R² =", base::round(stock_r2, 3), "\n")
base::cat("- RSE = ±$", base::round(stock_rse, 1), "per share\n")
base::cat("- Conclusion: High R-Squared (", base::round(stock_r2, 3), ") but ±$", base::round(stock_rse, 1), "\n")
base::cat("  errors still make this useless for trading!\n\n")

```




## The Key Insight

**R-Squared and absolute error size are partially independent:**

```{r}
#| label: tbl-key-insight
#| tbl-cap: "The Key Insight: R² vs Absolute Error"

key_insight <- base::data.frame(
  Scenario = base::c("Weather (°F)", "Weather (°K)", "Stock prices", "Manufacturing"),
  R_squared = base::c("0.85", "0.85", "0.85", "0.30"),
  Absolute_Error = base::c("±3°F", "±3°K", "±$50/share", "±0.1mm"),
  Practical_Value = base::c("Excellent!", "Same accuracy, different units", "Useless for trading", "Perfect for quality control")
)

flextable::flextable(key_insight) |>
  flextable::set_caption(caption = "The Key Insight: R² vs Absolute Error") |>
  flextable::align(j = "R_squared", align = "center", part = "body") |>
  flextable::bg(i = 1, bg = "#90EE90") |>
  flextable::bg(i = 2, bg = "#FFFF99") |>
  flextable::bg(i = 3, bg = "#FFB6C1") |>
  flextable::bg(i = 4, bg = "#90EE90") |>
  flextable::autofit() |>
  flextable::theme_vanilla()
```



## Why This Matters

**R² alone can be misleading because:**

1. **High R² + Large scale = Large absolute errors**
2. **Low R² + Small scale = Tiny absolute errors** 
3. **Practical utility depends on your error tolerance, not R²**

**Example:** Predicting tomorrow's temperature

- **Model A:** R² = 0.95, errors ±20°F → Terrible!
- **Model B:** R² = 0.60, errors ±5°F → Excellent!

The lesson: **Always check absolute error metrics (RSE, MAE) against your real-world tolerance, regardless of how impressive your R² looks!**
